---
output: html_document
editor_options: 
  chunk_output_type: console
---
Increasingly Powerful Tornadoes
===============================

James B. Elsner
---------------

This Rmd was written in support of _Increasingly powerful tornadoes in the United States_ by Elsner and Fricker.

Using R 3.5.1 Feather Spray. Unzip the `IncreasingPower` folder and open `IncreasingPowerFinal.Rmd` in RStudio.

Set multiple cores for Stan model. http://r.789695.n4.nabble.com/MC-CORES-and-mc-cores-for-parallel-package-td4727155.html
```{r}
Sys.setenv("MC_CORES"= 4)
options("mc.cores")
library(parallel)
options("mc.cores")
```

Get packages and load the Jun 7 data image. The `model` object from the image is saved from a 5-hr run of the hierarchical model written in Stan code using the **brms** package.
```{r}
library(tidyverse)
library(lubridate)
library(rgeos)
library(sf)
library(xtable)
load(file = "Jun7.RData")
```

Get the tornado data. Data are from the Storm Prediction Center (SPC) http://www.spc.noaa.gov/gis/svrgis/.
```{r, eval=FALSE}
if(!file.exists("torn")){ 
    download.file(url = "http://www.spc.noaa.gov/gis/svrgis/zipped/tornado.zip",
                  destfile = "tornado.zip")
    unzip("tornado.zip")
}
Tor.sfdf <- sf::st_read(dsn = "torn", 
                      layer = "torn", 
                      stringsAsFactors = FALSE)
```

The newest GIS shapefile contains missing geometries for more than 30% of the tornadoes. The number of missing geometries is highest after 1995. Instead use the csv file from https://www.spc.noaa.gov/wcm/#data  Use the start lon/lat and create a `sp` object then convert to `sf`. Set the coordinate reference system (crs) to ESPG 4326.
```{r}
Tor.df <- read.csv(file = "1950-2017_actual_tornadoes.csv")
Tor.spdf <- Tor.df
rm(Tor.df)
sp::coordinates(Tor.spdf) <- ~ slon + slat
Tor.sfdf <- st_as_sf(Tor.spdf)
st_crs(Tor.sfdf) <- 4326
```

Prepare the tornado data.
```{r}

Tor.sfdf <- Tor.sfdf %>%
  filter(yr >= 1994, mag != -9) %>%
  mutate(DateTime = as.POSIXct(paste(yr, mo, dy, time), format = "%Y%m%d%H:%M:%S"),
         Hour = hour(DateTime),
         Year = year(DateTime),
         Length = len * 1609.34,
         Length = ifelse(Length == 0, min(Length[Length > 0]), Length), #takes care of zero length
         Width = wid * .9144,
         Width = ifelse(Width == 0, min(Width[Width > 0]), Width), #takes care of zero width
         Width = ifelse(Year >= 1995, Width * pi/4, Width), #takes care of change: avg to max
         cas = inj + fat,
         AreaPath = Length * Width,
         Ma = factor(month.abb[mo], levels = month.abb[1:12])) %>%
  sf::st_sf()
```

Compute per-tornado energy dissipation. Energy dissipation is computed as in Fricker et al. (2017).
```{r}
perc <- c(1, 0, 0, 0, 0, 0, 
         .772, .228, 0, 0, 0, 0,
         .616, .268, .115, 0, 0, 0,
         .529, .271, .133, .067, 0, 0,
         .543, .238, .131, .056, .032, 0,
         .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
ef <- Tor.sfdf$mag + 1
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] <- midptW^3 %*% percM[ef[i], ]
Tor.sfdf$e3 <- EW3
Tor.sfdf$ED <- Tor.sfdf$e3 * Tor.sfdf$AreaPath
aM <- mean(Tor.sfdf$ED)/10^9
quantile(Tor.sfdf$ED, probs = c(.25, .5, .75))/10^9
gM <- exp(mean(log(Tor.sfdf$ED)))/10^9 # ED in units of gigawatts
aM; gM
scalingFactor <- aM/gM
```

Statistics of energy dissipation by EF rating.
```{r}
as.data.frame(Tor.sfdf) %>%
  group_by(mag) %>%
  dplyr::summarize(nT = n(),
                   Median = median(ED)/10^9,
                   Total = sum(ED)/10^9,
                   arithmeticMean = mean(ED)/10^9,
                   geometricMean = exp(mean(log(ED)))/10^9,
                   harmonicMean = 1/mean(1/ED)/10^9
                   ) %>%
xtable(., digits = 1)
```

Statistics of energy dissipation by hour of day.
```{r}
as.data.frame(Tor.sfdf) %>%
  group_by(Hour) %>%
  dplyr::summarize(nT = n(),
                   Median = median(ED)/10^9,
                   Total = sum(ED)/10^9,
                   arithmeticMean = mean(ED)/10^9,
                   geometricMean = exp(mean(log(ED)))/10^9,
                   q90 = quantile(ED, probs = .9)/10^9,
                   q75 = quantile(ED, probs = .75)/10^9
                   ) %>%
xtable(., digits = 2)
```

By date and by outbreak size. Showing that the percentage of strong and violent tornadoes increases with outbreak size is determined by the number of tornadoes per calendar day.
```{r}
levels <- c(0, 1, 3, 7, 15, 31, 63, Inf)
labels <- c("1", "2-3", "4-7", "8-15", "16-31", "32-63", ">63")
dfx <- as.data.frame(Tor.sfdf) %>%
  group_by(date) %>%
  summarize(nT = n(),
            nST = sum(mag >= 3),
            nVT = sum(mag >= 4)) %>%
  mutate(outbreakSize = cut(nT, levels, labels = labels)) %>%
  group_by(outbreakSize) %>%
  summarize(totalCases = n(),
    totalTornadoes = sum(nT),
    pST = sum(nST)/totalTornadoes * 100,
    pVT = sum(nVT)/totalTornadoes * 100)
dfx
```

Annual quantiles
```{r}
df <- as.data.frame(Tor.sfdf) %>%
  group_by(yr) %>%
  dplyr::summarize(nT = n(),
                   nC = sum(inj + fat),
                   avgED = mean(ED),
                   q25ED = quantile(ED, prob = .25),
                   q75ED = quantile(ED, prob = .75),
                   q90ED = quantile(ED, prob = .9),
                   q95ED = quantile(ED, prob = .95),
                   q50ED = quantile(ED, prob = .5))

p1 <- ggplot(df, aes(x = yr, y = q50ED/10^9)) +
  geom_errorbar(aes(ymin = q25ED/10^9, ymax = q75ED/10^9), color = "grey", width = .2) +
  geom_point() +
  geom_point(aes(x = yr, y = q90ED/10^9), data = df, color = "red") +
  scale_y_log10(breaks = c(.1, 1, 10, 100, 1000), labels = c(.1, 1, 10, 100, 1000), limits = c(.01, 1000)) +
  scale_x_continuous(breaks = seq(1995, 2015, 5)) +
  xlab("Year") + ylab("Energy Dissipation [GW]") +
  theme_minimal() 
```
#### Figure 1

Energy dissipation quantiles by year and month
```{r}
df <- as.data.frame(Tor.sfdf) %>%
  group_by(yr, mo) %>%
  dplyr::summarize(nT = n(),
                   nC = sum(inj + fat),
                   avgED = mean(ED),
                   q25ED = quantile(ED, prob = .25),
                   q75ED = quantile(ED, prob = .75),
                   q90ED = quantile(ED, prob = .9),
                   q95ED = quantile(ED, prob = .95),
                   q50ED = quantile(ED, prob = .5))

names(month.name) <- 1:12 # this gives the character vector names corresponding to `mo`.
p2 <- ggplot(df, aes(x = yr, y = q50ED/10^9)) +
  geom_errorbar(aes(ymin = q25ED/10^9, ymax = q75ED/10^9), color = "grey", width = .2) +
  geom_point() +
  geom_point(aes(x = yr, y = q90ED/10^9), data = df, color = "red", inherit.aes = FALSE) +
  scale_y_log10(breaks = c(.1, 1, 10, 100, 1000), labels = c(.1, 1, 10, 100, 1000)) +
  scale_x_continuous(breaks = seq(1995, 2015, 10)) +
  xlab("") + ylab("Energy Dissipation [GW]") +
  facet_wrap(~ mo, ncol = 12, labeller = as_labeller(month.name)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

df <- as.data.frame(Tor.sfdf) %>%
  group_by(yr) %>%
  dplyr::summarize(nT = n(),
                   nC = sum(inj + fat),
                   geometricMean = exp(mean(log(ED)))/10^9,
                   geometricSD = exp(sqrt(sum((log(ED/geometricMean))^2/(length(ED) - 1))))/10^9,
                   lb = geometricMean/geometricSD,
                   ub = geometricMean * geometricSD)

p2A <- ggplot(df, aes(x = yr, y = geometricMean)) +
  geom_errorbar(aes(ymin = lb, ymax = ub), color = "grey", width = .2) +
  geom_point() +
  scale_y_continuous() +
  scale_x_continuous(breaks = seq(1995, 2015, 10)) +
  xlab("") + ylab("Energy Dissipation [GW]") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

Energy quantiles by hour of day
```{r}
df <- as.data.frame(Tor.sfdf) %>%
  group_by(Hour) %>%
  dplyr::summarize(nT = n(),
                   nC = sum(inj + fat),
                   arithmeticED = mean(ED),
                   geometricMean = exp(mean(log(ED))),
                   q25ED = quantile(ED, prob = .25),
                   q75ED = quantile(ED, prob = .75),
                   q90ED = quantile(ED, prob = .9),
                   q95ED = quantile(ED, prob = .95),
                   q99ED = quantile(ED, prob = .99),
                   q50ED = quantile(ED, prob = .5),
                   maxED = max(ED))

p3 <- ggplot(df, aes(x = Hour, y = q50ED/10^9)) +
  geom_errorbar(aes(ymin = q25ED/10^9, ymax = q75ED/10^9), color = "grey", width = .2) +
  geom_point() +
  geom_point(aes(x = Hour, y = q90ED/10^9), data = df, color = "red") +
  scale_y_log10(breaks = c(.1, 1, 10, 100, 1000), labels = c(.1, 1, 10, 100, 1000), limits = c(.01, 1000)) +
  scale_x_continuous(breaks = 0:23) +
  xlab("Local Hour of Occurrence [+/- one hr]") + ylab("Energy Dissipation [GW]") +
  theme_minimal() +
  ggtitle("B")
```

Polar coordinate plots.
```{r}
labsHours <- c("Midnight", "2 am", "4 am", "6 am", "8 am", "10 am", "Noon", "2 pm", "4 pm", "6 pm", "8 pm", "10 pm")
c1 <- ggplot(df, aes(Hour, nT)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(0, 23, 2), 
                     labels = labsHours) +
  coord_polar(start = -0.12) +
  labs(x = "Time of Day", y = "Number of Tornadoes") +
  theme_minimal()
c2 <- ggplot(df, aes(Hour, geometricMean/10^9)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(0, 23, 2), 
                     labels = labsHours) +
  coord_polar(start = -0.12) +
  labs(x = "Time of Day", y = "Mean (geometric) Tornado Power (GW)") +
  theme_minimal()
```

Put multiple plots on the same page using **patchwork**.  Thanks to https://github.com/thomasp85/patchwork/ If `library(patchwork)` from Thomas Lin Pedersen returns an error. Install `ggplot2` from GitHub (Hadley Wickham). Restart your R session.
```{r}
#devtools::install_github("thomasp85/patchwork")
#devtools::install_github("hadley/ggplot2")
library(ggplot2)
library(patchwork)
c1 + c2
```

Repeat polar coordinate plot with monthly means.
```{r}
df %>%
  group_by(mo) %>%
  summarize(arthmeticMean = mean(ED)/10^9,
            geometricMean = exp(mean(log(ED)))/10^9,
            geometricSD = exp(sqrt(sum((log(ED/10^9/geometricMean))^2/(length(ED/10^9) - 1)))),
            q25ED = quantile(ED, prob = .25)/10^9,
            q75ED = quantile(ED, prob = .75)/10^9,
            q50ED = quantile(ED, prob = .5)/10^9) %>%
ggplot(., aes(x = mo, y = q50ED)) +
  geom_bar(stat = "identity", width = 1) +
#  geom_errorbar(aes(ymin = q25ED, 
#                    ymax = q75ED),
#                    width = .2) +
  scale_x_continuous(breaks = seq(1, 12, 1), labels = month.abb) +
  scale_y_continuous(limits = c(0, NA)) +
  coord_polar(start = 0) +
  labs(x = "Month", y = "Average (geometric) Tornado Power (GW)") +
  theme_minimal()
```

Compare with monthly and hourly distributions of casualties.
```{r}
df <- as.data.frame(Tor.sfdf) %>%
  filter(cas > 0) %>%
  group_by(mo) %>%
  summarize(nT = n(),
            nCas = sum(cas),
            perTorCas = nCas/nT)
ggplot(df, aes(x = mo, y = perTorCas)) +
  geom_bar(stat = "identity", width = 1) +
  scale_x_continuous(breaks = seq(1, 12, 1), labels = month.abb) +
  coord_polar(start = 0) +
  labs(x = "Month", y = "Number of Casualties Per\n Casualty-Producing Tornado") +
  theme_minimal()

df <- as.data.frame(Tor.sfdf) %>%
  filter(cas > 0) %>%
  group_by(Hour) %>%
  summarize(nT = n(),
            nCas = sum(cas),
            perTorCas = nCas/nT)
ggplot(df, aes(x = Hour, y = perTorCas)) +
  geom_bar(stat = "identity", width = 1) +
  scale_x_continuous(breaks = seq(0, 23, 2), labels = labsHours) +
  coord_polar(start = -.12) +
  labs(x = "Time of Day", y = "Casualty Rate Per\n Casualty-Producing Tornado") +
  theme_minimal()
```

Try to smooth these counts.
```{r}
y <- df$perTorCas
y <- c(y, y, y)
x.mid <- 1:24; offset <- 24
y.smooth <- lowess(y, f = 1/4)
df$ys <- y.smooth$y[x.mid + offset]

ggplot(df, aes(x = Hour, y = perTorCas)) +
  geom_bar(stat = "identity", width = 1) +
  geom_line(aes(y = ys), color = "red") +
  scale_x_continuous(breaks = seq(0, 23, 2), labels = labsHours) +
  coord_polar(start = -.12) +
  labs(x = "Local Time of Day", y = "Casualty Rate Per\n Casualty-Producing Tornado") +
  theme_minimal()
```

Putting multiple plots together in a patchwork.
```{r}
(p1 + p3) / p2
```

Get the ENSO variable http://www.esrl.noaa.gov/psd/data/correlation/censo.data. Here we use the bivariate ENSO. Join to per tornado data frame with the `left_join()` function. We also get the western Caribbean SST (WCA) as used in the seasonal prediction model. 

SST bbox: 10, 35, -97, -70  https://www.esrl.noaa.gov/psd/cgi-bin/data/timeseries/timeseries1.pl
```{r}
ENSO <- read.table("ENSO.txt", header = TRUE) %>%
  filter(Year >= 1994)
ENSO.df <- reshape2::melt(ENSO, id.vars = "Year")
names(ENSO.df) <- c("yr", "mo.abb", "ENSO")

SST <- read.table("SST.txt", header = TRUE) %>%
  filter(Year >= 1994)
SST.df <- reshape2::melt(SST, id.vars = "Year")
names(SST.df) <- c("yr", "mo.abb", "SST")

df <- as.data.frame(Tor.sfdf) %>%
  mutate(mo.abb = month.abb[mo]) %>%
  left_join(ENSO.df)
df <- left_join(df, SST.df)
df$After2007 <- df$yr >= 2007
```

Change point in 2007. Use of the Enhanced Fujita Scale.

The geometric standard deviation describes the spread in the set of energy dissipation values as a (dimensionless) multiplicative factor relative to the geometric mean. The one standard deviation range about the mean is obtained by dividing and multipling the geometric mean by the geometric standard deviation.
$$\sigma_g = \exp \left( \sqrt{ \sum_{i=1}^n \left( \ln { A_i \over \mu_g } \right)^2 \over n-1 } \right)$$

```{r}
p4 <- df %>%
  group_by(After2007) %>%
  summarize(arthmeticMean = mean(ED)/10^9,
            geometricMean = exp(mean(log(ED)))/10^9,
            geometricSD = exp(sqrt(sum((log(ED/geometricMean))^2/(length(ED) - 1))))/10^9,
            lb = geometricMean/geometricSD,
            ub = geometricMean * geometricSD,
            harmonicMean = 1/mean(1/ED)/10^9) %>%
  ggplot(., aes(x = After2007, y = geometricMean)) +
  geom_errorbar(aes(ymin = lb, ymax = ub), col = "grey", width = 0, size = 2) +
  geom_point() +
  xlab("") + ylab("Energy Dissipation [GW]") +
  scale_x_discrete(labels = c("F Scale", "EF Scale")) +
  scale_y_continuous(limits = c(0, NA)) +
  ggtitle("A") +
  theme_minimal()
```

Repeat for path length and width
```{r}
df %>%
  group_by(After2007) %>%
  summarize(arithmeticMean = mean(Width),
            geometricMean = exp(mean(log(Width))),
            geometricSD = exp(sqrt(sum((log(Width/geometricMean))^2/(length(Width) - 1)))),
            lb = geometricMean/geometricSD,
            ub = geometricMean * geometricSD) %>%
  ggplot(., aes(x = After2007, y = geometricMean)) +
  geom_errorbar(aes(ymin = lb, ymax = ub), col = "grey", width = 0, size = 2) +
  geom_point() +
  xlab("") + ylab("Length [m]") +
  scale_x_discrete(labels = c("F Scale", "EF Scale")) +
  scale_y_continuous(limits = c(0, NA)) +
  theme_minimal()

df %>%
  group_by(After2007) %>%
  summarize(nT = n(),
            nT345 = sum(mag >=3)/nT)
```

Joy plots
```{r}
library(ggridges)
S1 <- ggplot(df, aes(x = Length/1000, y = Year, group = Year, height = ..density..)) +
     geom_density_ridges(scale = 5, color = "white") + 
     scale_x_log10(breaks = c(.1, 1, 10, 100), labels = c(".1", "1", "10", "100")) +
     scale_y_reverse(breaks = seq(1995, 2015, 5)) +
     ylab("") + xlab("Path Length (km)")

S2 <- ggplot(df, aes(x = Width, y = Year, group = Year, height = ..density..)) +
     geom_density_ridges(scale = 5, color = "white") + 
     scale_x_log10(limits = c(1, NA), breaks = c(.1, 1, 10, 100), labels = c(".1", "1", "10", "100")) +
     scale_y_reverse(breaks = seq(1995, 2015, 5)) +
     ylab("") + xlab("Path Width (m)")
S1 + S2
```
#### Figure S1 
Path widths less than 1 meter are not plotted.

Monthly and hourly
```{r}
p5 <- df %>%
  group_by(Ma) %>%
  summarize(arthmeticMean = mean(ED)/10^9,
            geometricMean = exp(mean(log(ED)))/10^9,
            geometricSD = exp(sqrt(sum((log(ED/geometricMean))^2/(length(ED) - 1))))/10^9,
            lb = geometricMean/geometricSD,
            ub = geometricMean * geometricSD,
            harmonicMean = 1/mean(1/ED)/10^9) %>%
ggplot(., aes(x = Ma, y = geometricMean)) +
  geom_errorbar(aes(ymin = lb, ymax = ub), col = "grey", width = 0, size = 2) +
  geom_point() +
  xlab("") + ylab("Energy Dissipation [GW]") +
  scale_y_continuous(limits = c(0, NA)) +
  ggtitle("C") +
  theme_minimal()

labelsHour <- c("Midnight", "3am", "6am", "9am", "Noon", "3pm", "6pm", "9pm")
p6 <- df %>%
  group_by(Hour) %>%
  summarize(arthmeticMean = mean(ED)/10^9,
            geometricMean = exp(mean(log(ED)))/10^9,
            geometricSD = exp(sqrt(sum((log(ED/geometricMean))^2/(length(ED) - 1))))/10^9,
            lb = geometricMean/geometricSD,
            ub = geometricMean * geometricSD,
            harmonicMean = 1/mean(1/ED)/10^9) %>%
ggplot(., aes(x = Hour, y = geometricMean)) +
  geom_errorbar(aes(ymin = lb, ymax = ub), col = "grey", width = 0, size = 2) +
  geom_point() +
#  xlab("Local Hour of Occurrence [+/- one hr]") 
  xlab("") + ylab("Energy Dissipation [GW]") +
  scale_x_continuous(breaks = seq(0, 23, by = 3), 
                     labels = labelsHour) +
  scale_y_continuous(limits = c(0, NA)) +
  ggtitle("D") +
  theme_minimal()

p7 <- df %>%
  group_by(ENSO > 0) %>%
  summarize(arthmeticMean = mean(ED)/10^9,
            geometricMean = exp(mean(log(ED)))/10^9,
            geometricSD = exp(sqrt(sum((log(ED/geometricMean))^2/(length(ED) - 1))))/10^9,
            lb = geometricMean/geometricSD,
            ub = geometricMean * geometricSD,
            harmonicMean = 1/mean(1/ED)/10^9) %>%
ggplot(., aes(x = `ENSO > 0`, y = geometricMean)) +
  geom_errorbar(aes(ymin = lb, ymax = ub), col = "grey", width = 0, size = 2) +
  geom_point() +
  xlab("") + ylab("Energy Dissipation [GW]") +
  scale_x_discrete(labels = c("La Nina", "El Nino")) +
  scale_y_continuous(limits = c(0, NA)) +
  ggtitle("B") +
  theme_minimal()

(p4 + p7 + p5 + plot_layout(ncol = 3, widths = c(1, 1, 3))) / p6
```
#### Figure 2

Path length by EF rating and month. Hypothesis: violent tornadoes have much shorter paths during May than during November.
```{r}
df %>%
  filter(mag >= 3) %>%
  group_by(mo) %>%
  summarize(medianPathLength = median(Length))

df %>%
  group_by(mo) %>%
  summarize(nT = n(),
            nT3 = sum(mag >= 3),
            nT4 = sum(mag >= 4),
            nT5 = sum(mag == 5),
            mPL3 = median(Length[mag >= 3]),
            mPL4 = median(Length[mag >= 4]),
            mPL5 = median(Length[mag == 5]))
```

### Models

Start with a log-linear mixed model using the `lmer()` function from the **lme4** package. The response variable is the natural logarithm of energy dissipation.
```{r}
library(lme4)
model0 <- lmer(log(ED) ~  ENSO + After2007 + I(yr - 2002) + (1|mo) + (1|Hour), data = df)
model0a <- lmer(log(ED) ~ ENSO + SST + After2007 + I(yr - 2002) + (1|mo) + (1|Hour), data = df)

summary(model0a)
ranef(model0a)
```

All Bayesian models were created in Stan computational framework (http://mc-stan.org/) accessed with **brms** package (Burkner 2017). To improve convergence and guard against overfitting, we specified mildly informative conservative priors.
```{r}
library(brms)
```

The most important reason to use `control` is to decrease (or preferably) eliminate the number of divergent transitions that cause a bias in the posterior samples. With a warning `There were divergent transitions after warmup` you should increase the `adapt_delta`. For example `control = list(adapt_delta = .95)`. The default is a  value of .8. Increasing the value slows down the sampler.
```{r}
#formula <- ED | trunc(lb = 444000) ~ ENSO + After2007 + I(yr - 2002) + s(mo) + s(Hour)  # Model in the May2.RData file
formula <- ED | trunc(lb = 444000) ~ ENSO + After2007 + I(yr - 2002) + (1|mo) + (1|Hour) # Model as in the lmer specification
family <- brms::lognormal()
```

The model formula is a truncated log-normal distribution to reflect a threshold energy below which tornadoes have never occurred. We set this threshold at 4.440 GW just below the 4.445 GW weakest tornado in the record.

Laptop errors with `cores` > 1?
```{r, eval=FALSE}
df2 <- df %>%
  filter(mag >= 0)
model <- brm(family = family,
             formula = formula,
             data = df2,
             prior = set_prior('normal(0, 1)'),
#             warmup = 1000, iter = 5000, chains = 4, cores = 4,
             warmup = 200, iter = 2200, chains = 4, cores = 4,
             inits = "0",
             control = list(adapt_delta = .8))
```


```{r}
save.image(file = "Jun7.RData")
```

There were 390 transitions after warmup that exceeded the maximum treedepth. Increase max_treedepth above 10. See
http://mc-stan.org/misc/warnings.html#maximum-treedepth-exceededExamine the pairs() plot to diagnose sampling problems

```{r}
summary(model)
```

The wiggly parts of the spline basis are treated as a random effects and their associated variance parameter `sigma` controls the degree of wiggliness of the fitted spline. The perfectly smooth parts of the basis are treated as a fixed effect (`smo_1`).

```{r}
coefTable <- as.data.frame(summary(model)$fixed) %>%
  mutate(lb = exp(`l-95% CI`),
         ub = exp(`u-95% CI`),
         mi = exp(`Estimate`),
         id = 1:6) %>%
  filter(id > 1, id < 5)

ggplot(coefTable, aes(x = id, y = mi)) +
  geom_hline(yintercept = 1, color = "grey", linetype = "solid") +
  geom_hline(yintercept = c(.75, 1.25, 1.5, 1.75), color = "grey", linetype = "dashed") +
  geom_point() +  
  geom_errorbar(aes(x = id, ymin = lb, ymax = ub), col = "red", width = 0, size = 2) +
  geom_point() +
  scale_x_reverse(breaks = 2:4, labels = c("ENSO", "EF Rating", "Trend")) +
  scale_y_continuous(limits = c(.75, 1.75)) +
  ylab("Multiplicative Change") + xlab("") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  coord_flip() 
```

The **sjstats** package for model summaries on models fit using the **brms** package.
https://www.r-bloggers.com/r-functions-for-bayesian-model-statistics-and-summaries-rstats-stan-brms/

```{r}
library(sjstats)
hdi(model, type = "fixed")
```

```{r}
tidy_stan(model)
```


```{r, eval=FALSE}
pp_check(model, type = "stat")
#pp_check(model, type = "ecdf_overlay")
```

The distribution of the posterior mean energy dissipation (on a log scale) against the average. It looks good.

The `posterior_predict()` function generates samples of ED. Compare the distribution of ED statistics (mean, max) with actual ED.
```{r}
yrep <- posterior_predict(model, nsamples = 8000)
df.yrep <- as.data.frame(yrep)
df.out <- reshape2::melt(df.yrep) %>%
  group_by(variable) %>%
  summarize(mx = max(value),
            mn = exp(mean(log(value))))
```

```{r}
p8 <- ggplot(df.out, aes(mn/10^9)) + 
  geom_histogram(fill = "red", bins = 21, color = "white") +
#  geom_freqpoly() +
  scale_x_log10(breaks = c(1, 5, 10, 20, 40, 80)) +
  geom_vline(xintercept = exp(mean(log(df2$ED)))/10^9, color = "black", size = 1) +
  ylab("Posterior Frequency") +
  xlab("Average Per-Tornado Power [GW]") +
#  ggtitle("A") +
  theme_minimal()

p9 <- ggplot(df.out, aes(mx/10^9)) + 
  scale_x_log10(breaks = c(10000, 100000, 1000000), labels = c("10,000", "100,000", "1,000,000")) +
  geom_histogram(fill = "red", bins = 21, color = "white") +
  geom_vline(xintercept = max(df2$ED)/10^9, color = "black", size = 1) +
  ylab("Posterior Frequency") +
  xlab("Maximum Per-Tornado Power [GW]") +
#  ggtitle("B") +
  theme_minimal()

p8 + p9
```

Conditioning plots
```{r, eval=FALSE}
plot(marginal_effects(model))
```

The conditionals are pre-specified.
```{r}
EFmarginal <- marginal_effects(model, method = "fitted")$After2007

p10 <- ggplot(EFmarginal, aes(x = After2007, y = estimate__/10^9)) +
  geom_errorbar(aes(ymin = lower__/10^9 , ymax = upper__/10^9), col = "red", width = 0, size = 2) +
  geom_point() +
#  scale_y_log10() +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_discrete(labels = c("F Scale", "EF Scale")) +
  ylab("Conditional Energy Dissipation [GW]") + xlab("") +
  ggtitle("A") +
  theme_minimal()

ENSOmarginal <- marginal_effects(model)$ENSO

p11 <- ggplot(ENSOmarginal, aes(x = ENSO, y = estimate__/10^9)) +
  geom_ribbon(aes(ymin = lower__/10^9 , ymax = upper__/10^9), alpha = 1, col = "red", fill = "red") +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_continuous() +
  geom_line(size = 1.2) +
  ylab("Conditional Energy Dissipation [GW]") + xlab("ENSO [s.d.]") +
  ggtitle("B") +
  theme_minimal()

Monthmarginal <- marginal_effects(model)$mo %>%
  filter(mo %in% 1:12)

p12 <- ggplot(Monthmarginal, aes(x = mo, y = estimate__/10^9)) +
  geom_errorbar(aes(ymin = lower__/10^9, ymax = upper__/10^9), col = "red", width = 0, size = 2) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  geom_point() +
  xlab("") + ylab("Conditional Power [GW]") +
#  ggtitle("C") + 
  theme_minimal()

Hourmarginal <- marginal_effects(model)$Hour
HourmarginalI <- data.frame(Hour = 0:23,
                            estimate__ = approx(x = Hourmarginal$Hour, y = Hourmarginal$estimate__, xout = 0:23)$y,
                            lower__ = approx(x = Hourmarginal$Hour, y = Hourmarginal$lower__, xout = 0:23)$y,
                            upper__ = approx(x = Hourmarginal$Hour, y = Hourmarginal$upper__, xout = 0:23)$y)

p13 <-ggplot(HourmarginalI, aes(x = Hour, y = estimate__/10^9)) +
  geom_errorbar(aes(ymin = lower__/10^9, ymax = upper__/10^9), col = "red", width = 0, size = 2) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(0, 23, by = 3), labels = labelsHour) +
  geom_point() +
  xlab("") + ylab("Conditional Energy Dissipation [GW]") +
  ggtitle("D") + 
  theme_minimal()

(p10 + p11 + p12 + plot_layout(ncol = 3, widths = c(1, 1, 3))) / p13
```

Trend. With `method = "fitted"` the graph is computed as the arithmetic average ED over all the outcomes in the event space (e.g., yr = 1994, Hour = 15, mo = 6, etc). With `method = "predict"` the graph is the predicted geometric mean for specific values of the covariaties. We divide the fitted by the scaling factor.
```{r}
TrendmarginalF <- marginal_effects(model, method = "fitted")$yr
TrendmarginalP <- marginal_effects(model, method = "predict")$yr

ggplot(TrendmarginalF, aes(x = yr, y = estimate__/10^9/scalingFactor)) +
  geom_ribbon(aes(ymin = lower__/10^9/scalingFactor, ymax = upper__/10^9/scalingFactor), 
              alpha = 1, col = "red", fill = "red") +
  scale_y_continuous(limits = c(0, NA)) +
#  scale_y_log10() +
  geom_line(size = 1.2) +
  ylab("Conditional Power [GW]") + xlab("") +
  theme_minimal()
```

Specify the conditions.
```{r}
conditions <- data.frame(mo = 5, ENSO = 0, After2007 = TRUE, Hour = 16)
TrendmarginalF <- marginal_effects(model, conditions = conditions, method = "fitted")$yr

ggplot(TrendmarginalF, aes(x = yr, y = estimate__/10^9)) +
  geom_ribbon(aes(ymin = lower__/10^9 , ymax = upper__/10^9), alpha = 1, col = "red", fill = "red") +
  scale_y_continuous() +
  scale_x_continuous(breaks = seq(1995, 2015, 5)) +
  geom_line() +
  ylab("Predicted Energy Dissipation [GW]") + xlab("") +
  theme_minimal()

conditions <- data.frame(mo = 1:12, ENSO = 0, After2007 = TRUE, Hour = 16)
TrendmarginalF <- marginal_effects(model, conditions = conditions, method = "fitted")$yr

p2 + geom_ribbon(aes(x = yr, ymin = lower__/10^9/scalingFactor , ymax = upper__/10^9/scalingFactor), 
                 data = TrendmarginalF, alpha = .4, fill = "red", inherit.aes = FALSE) +
     geom_line(aes(x = yr, y = estimate__/10^9/scalingFactor), data = TrendmarginalF, col = "black", 
               inherit.aes = FALSE)
```
#### Figure 3

```{r}
conditions <- data.frame(mo = 1:12, ENSO = 0, After2007 = TRUE, Hour = 10)
Trendmarginal2 <- marginal_effects(model, effects = c("yr"), conditions = conditions, method = "fitted")$yr
Trendmarginal2p <- marginal_effects(model, conditions = conditions, method = "predict", probs = c(.25, .75))$yr

ggplot(Trendmarginal2p, aes(x = yr, y = estimate__/10^9)) +
#  geom_ribbon(aes(ymin = lower__/10^9 , ymax = upper__/10^9), alpha = 1, col = "red", fill = "red") +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(1995, 2015, 10)) +
  geom_line() +
  ylab("Conditional Energy Dissipation [GW]") + xlab("") +
  facet_wrap(~ mo, ncol = 12, labeller = as_labeller(month.name)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

conditions <- data.frame(mo = 1:12, ENSO = rep(c(2, -2), each = 12), After2007 = rep(c(FALSE, TRUE), each = 24))
Trendmarginal2 <- marginal_effects(model, conditions = conditions)$yr

ggplot(Trendmarginal2, aes(x = yr, y = estimate__/10^9)) +
  geom_ribbon(aes(ymin = lower__/10^9 , ymax = upper__/10^9), alpha = 1, col = "red", fill = "red") +
  scale_y_continuous(limits = c(0, NA)) +
  geom_line() +
  ylab("Conditional Energy Dissipation [GW]") + xlab("Year") +
  facet_wrap(~ ENSO + mo, ncol = 3) +
  theme_minimal()

df2 <- Trendmarginal2 %>%
  filter(yr == 2016, After2007 == TRUE)

ggplot(df2, aes(x = mo, y = estimate__/10^9, color = as.factor(ENSO))) +
  geom_errorbar(aes(x = mo, ymin = lower__/10^9, ymax = upper__/10^9, color = as.factor(ENSO)), 
                width = 0, size = 1, position = position_dodge(width = .5)) +
  geom_point(aes(x = mo, y = estimate__/10^9, color = as.factor(ENSO)), position = position_dodge(width = .5)) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_color_discrete(name = "", 
                       breaks=c(-2, 2),
                       labels=c("La Nina", "El Nino")) +
  xlab("") + ylab("Conditional Energy Dissipation [GW]") +
  theme_minimal() 

conditions <- data.frame(mo = 2, ENSO = 0, After2007 = TRUE, Hour = 18)
Trendmarginal3 <- marginal_effects(model, conditions = conditions)$yr

ggplot(Trendmarginal3, aes(x = yr, y = estimate__/10^9)) +
  geom_ribbon(aes(ymin = lower__/10^9 , ymax = upper__/10^9), alpha = 1, col = "red", fill = "red") +
  scale_y_continuous(limits = c(0, NA)) +
  geom_line() +
  ylab("Conditional Energy Dissipation [GW]") + xlab("Year") +
  theme_minimal()
```

Compare with Stan model fixed effects with the same fixed effects from an INLA model.
```{r}
#source("http://www.math.ntnu.no/inla/givemeINLA.R")
library('INLA')
```

```{r}
df2 <- df %>%
  filter(mag >= 0)
formula <- ED ~ ENSO + After2007 + I(yr - 2002) +
  f(Ma, model = "seasonal", season.length = 12, param = c(1, .1)) +
  f(Hour, model = "seasonal", season.length = 24, param = c(1, .1))
family <- "lognormal"
startTime <- Sys.time()
modeli <- inla(formula = formula, 
               family = family,
               data = df2,
               control.compute = list(config = TRUE))
Sys.time() - startTime
summary(modeli)
```

Compare fixed effects Stan vs INLA.
```{r}
coefTable2 <- as.data.frame(modeli$summary.fixed) %>%
  mutate(lb2 = exp(`0.025quant`),
         ub2 = exp(`0.975quant`),
         mi2 = exp(`mean`),
         id = seq(1.25, 4.25, by = 1)) %>%
  filter(id > 1.25)

ggplot(coefTable, aes(x = id, y = mi)) +
  geom_hline(yintercept = 1, color = "grey", linetype = "dashed") +
  geom_point() +  
  geom_errorbar(aes(x = id, ymin = lb, ymax = ub), col = "red", width = 0, size = 2) +
  geom_point() +
  geom_point(aes(x = id, y = mi2), data = coefTable2) +
  geom_errorbar(aes(x = id, y = mi2, ymin = lb2, ymax = ub2), data = coefTable2, col = "red", width = 0, size = 2) +
  geom_point(aes(x = id, y = mi2), data = coefTable2) +
  scale_x_reverse(breaks = c(2.125, 3.125, 4.125), labels = c("ENSO", "EF Rating", "Trend")) +
  scale_y_continuous(breaks = seq(.8, 1.6, .2), labels = seq(.8, 1.6, .2), limits = c(.75, 1.7)) +
  ylab("Multiplicative Change") + xlab("") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  coord_flip()
```
